---
title: "Data"
---

```{r}
#| label: data-setup
#| include: false
suppressPackageStartupMessages({
  library(dplyr)
  library(knitr)
  library(kableExtra)
})
load("data/HNS_2022_clean.RData")
```

## What is this dataset?

The dataset captures **hazardous cargo (HNS, Hazardous and Noxious Substances) handled at Turkish ports during 2022**. It was published by the **Directorate General for Maritime Affairs** (T.C. Ulaştırma ve Altyapı Bakanlığı – Denizcilik Genel Müdürlüğü) and shared with us in the form of a pivot-table workbook prepared by the directorate's intern team.

Each leaf record describes a single cargo movement at the resolution of:

-   the **regional port authority** (e.g., *Aliağa Bölge Liman Başkanlığı*),
-   the **facility** that handled the cargo (e.g., *Egegaz Aliağa LNG Terminal*),
-   the **receiver / final consignee company**,
-   the **cargo amount in metric tons**.

### Why this dataset?

We chose it for three reasons:

1.  **It's real, operational, and Turkish.** No simulated toy data, the patterns we surface are about the actual industrial geography of Türkiye's coastlines.
2.  **It supports policy questions.** The Maritime Affairs office has flagged firms handling more than 20,000 tons of hazardous cargo as candidates for upcoming tax/regulatory measures. The dataset lets us put numbers on that conversation.
3.  **It's multi-dimensional.** Company × facility × port authority gives us room for descriptive, geographic, and threshold-based analyses without leaving a single source.

## Source

> **Directorate General for Maritime Affairs.** *2022 Hazardous Cargo Handling Records.* Internal compilation, accessed via project supervisor.

The original Excel workbook is bundled with this site as `data/HNS_2022.xlsx`, alongside the cleaned versions described below.

## Preparation pipeline

The raw workbook is a pivot-table export with three nested categorical levels (port authority → facility → company) and built-in subtotal rows ("Toplam …"). To analyse it we need a flat, tidy table.

The full preprocessing script lives at `R/prepare_data.R`. In short, it:

1.  Reads the workbook with `readxl::read_excel()` and assigns clean column names: `port_authority`, `facility`, `company`, `cargo_mt`.
2.  **Drops every subtotal row** (`Toplam …`) at all three hierarchy levels.
3.  Drops rows where the company or tonnage is missing.
4.  Renames the `(boş)` placeholder facility (used at Güllük) to `Unspecified facility`.
5.  Builds a **`company_clean`** field, uppercased, ASCII-folded, punctuation-stripped, for grouping near-duplicate company names.
6.  Extracts a **`port_authority_short`** field by stripping the boilerplate "Bölge Liman Başkanlığı" suffix.
7.  Joins a **province + ISO-2 code** lookup so we can map each port authority onto a Turkish province for the choropleth.

Re-run the script any time the source workbook changes:

``` r
source("R/prepare_data.R")
```

It writes:

-   `data/HNS_2022_clean.RData`, the canonical R object.
-   `data/HNS_2022_clean.csv`, the same table in CSV (UTF-8) for inspection / sharing.

## Download the cleaned data

::: callout-tip
The cleaned R dataset is available here: [**HNS_2022_clean.RData**](data/HNS_2022_clean.RData) (also available as [HNS_2022_clean.csv](data/HNS_2022_clean.csv)).
:::

Loading it back is a one-liner:

``` r
load("data/HNS_2022_clean.RData")  # creates a data frame called `hns2022`
```

## Schema at a glance

| Column | Type | Description |
|------------------------|------------------------|------------------------|
| `port_authority` | character | Full port-authority name (e.g., *Aliağa Bölge Liman Başkanlığı*) |
| `port_authority_short` | character | Short form (e.g., *Aliağa*), used for joins |
| `facility` | character | Facility / terminal that handled the cargo |
| `company` | character | Receiver / consignee, as written in the source |
| `company_clean` | character | Normalised company name for grouping near-duplicates |
| `cargo_mt` | numeric | Cargo amount, in metric tons |
| `province` | character | Host province of the port authority |
| `iso2` | character | ISO 3166-2 code for the province (e.g., `TR-35`) |

## A peek at the data

```{r}
#| label: tbl-head-rows
#| tbl-cap: "First 10 rows of the cleaned dataset."
#| code-fold: true
hns2022 |>
  dplyr::select(port_authority_short, facility, company, cargo_mt, province) |>
  head(10) |>
  kable(
    col.names = c("Port authority", "Facility", "Company", "Cargo (m. ton)", "Province"),
    format.args = list(big.mark = ",")
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```
